Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-07-08 main.tex
Philippe ajout deux exemples

\begin{example}
Let us develop the example of comparison between music played by a performer,
represented as a sequence $s \in \Sigma^*$ of events in the MIDI alphabet $\Sigma$,
and  a music score represented as a sequence $t \in \Delta^*$
in the CWMN alphabet $\Delta$. We build a small weighted transducer model
 with two states $q_0$ and $q_1$ that
calculates the distance between $s$ and $t$.

@@

@@

\begin{example}
structured words with timed symbols...
\florent{todo example VPA}
intro language of music notation? (markup = time division, leaves = events etc)
\end{example}

@@

\item[--] a \SWVPA $A$ over $\Delta$, $\Semiring$, and $\bar\Phi$,
defining a measure $A: \Delta^* \to \Semiring$,
%a \SWVPA $A$ over $\Delta$, and $\Semiring$, defining a series of nested words
%      $A : \Delta^* \to \Semiring$,
\item[--] an input word $s \in \Sigma^*$.
\end{description}
%
For all $u \in \Sigma^*$ and $t \in \Delta^*$, let
\(d(u, t) = T\bigl(u, t|_\Deltai \bigr)\),
where $t|_\Deltai \in \Deltai^*$ is the projection of $t$ onto~$\Deltai$,
obtained from $t$ by removing all symbols in $\Delta \setminus \Deltai$.
%\end{equation}
%
\noindent
\emph{Symbolic weighted parsing} is the problem,
given the above input,
to find $t \in \Delta^*$ %nested word $t \in \Delta$
minimizing \( d(s, t) \otimes A(t)\)
\wrt $\leq_\oplus$,
\ie s.t. %such that: %Hence, it is the problem of finding
%
\begin{equation}\label{eq:distance-lang}
d(s, t) \otimes A(t) = \displaystyle\bigoplus_{t' \in \Delta^*} d(s, t') \otimes A(t')
\end{equation}
%
Following the terminology of~\cite{Mohri03EDWA},
$\SW$-parsing is the problem of computing
the distance~\eqref{eq:distance-lang} between the input~$s$ and the output weighted language of~$A$,
and returning a witness $t$.
%
%\subsection{Computation}
%
\begin{proposition}
The problem of Symbolic Weighted  parsing
can be solved in PTIME in the size of the input \SWT $T$, \SWVPA $A$
and input word $s$,
and the computation time of the functions and operators of the label theory.
\end{proposition}
%
\begin{proof} (sketch)
We follow a \emph{Bar-Hillel} construction, for parsing by intersection.
%
\noindent
Let us first extend the \SWT $T$ over $\Sigma$, $\Deltai$
into a \SWT $T'$ over $\Sigma$ and $\Delta$
(and the same semiring and label theory $\Semiring$ and $\bar\Phi$),
such that for all $u \in \Sigma^*$, and $t \in {\Delta}^*$,
$T'(u, u) = T(u, t|_{\Deltai})$.
%
The transducer $T'$ simply skips every symbol
$b \in {\Delta} \setminus \Deltai$,
by the addition to $T$,
of new transitions of the form $\wei_{01}(q, \varepsilon, b, q')$.
%
\noindent
Then, using Corolary~\ref{cor:epsilon},
we construct from the input word $s \in \Sigma^*$ and $T'$
a \SWA $B_{s, T'}$,
such that for all $t \in \Delta^*$, $B_{s, T'}(t) = d(s, t)$.
%
%This automaton is such that for all $t \in \Delta^*$,
%\[
%   A_{T', s}\bigl(\lin(t)\bigr)
% = T'\bigl(s, \lin(t)\bigr)
% = T'\bigl(s, \lin(t)|_{\Deltai}\bigr)
% = d(s, t).
%\]
%
\noindent
Next, %we convert the input \SWTA $A$ over $\Delta$
%into a \SWVPA $A'$ over $\hat\Delta$, using Proposition~\ref{lem:SWTA}, and
we compute the \SWVPA $B_{s, T'} \otimes A$,
using Proposition~\ref{prop:SWVPA-product}.
%
\noindent
It remains to compute a best nested-word $t \in {\Delta}^*$
using the best-search procedure of Proposition~\ref{th:best-search}.
%and convert it into a best tree in $\T_\Delta$ in order to solve SW parsing
%for~$T$, $A$ and~$s$.
%\qed
\end{proof}
The $\SW$-parsing generalizes
the problem of searching the best derivation (AST) of a weighted CF-grammar $G$
that yields a given input word $w$.
The latter problem, sometimes called~\emph{weighted parsing},
(see \eg~\cite{Goodman99SemiringParsing}
 and~\cite{MorbitzVogler19weighted-parsing} for general weighted parsing frameworks)
corresponds to $\SW$-parsing in the case of finite alphabets,
a transducer $T$ computing the identity and some \SWVPA~$A$
obtained from $G$. %the weighted CF grammar.
%
%Indeed, it corresponds to the case where $T$
%accepts only the pairs $\<s, t>$ such that
%$s$ is the projection of $t$ on $\Deltai$.
%This can be done with a single state $q$ and
%with transition rules of the form:
%\begin{description}
%\item[] $\wei(q, \varepsilon, a, q) = \one$ for all $a \in \Deltac \cup \Deltar$,
%\item[] $\wei(q, a, a, q) = \one$ for all $a \in \Deltai$,
%\item[] $\wei(q, a, b, q) = \zero$ for all $a, b \in \Deltai$, $a \neq b$.
%\end{description}
%
Indeed, the \emph{depth-first} traversal of an AST $\tau$
yields a well-parenthesised word $\lin(\tau)$ over an alphabet
$\Delta = \Deltai \uplus \Deltac \uplus \Deltar$,
assuming \eg that $\Deltai$ contains the symbols labelling the leaves of $\tau$ (symbols of rank $0$),
and $\Deltac$ and $\Deltar$ contain respectively one left and right parenthesis
$\ccall{b}$ and $\creturn{b}$ for each symbol $b$ labelling inner nodes of $\tau$ (symbols of rank $>0$).
%
%With this representation, the projection $\lin(t)|_\Deltai$ is the
%sequence of leaves of $\tau$. %, enumerated in a \emph{dfs}-traversal.
%
We show in Appendix~\ref{sec:trees} how to construct 
%($\SW$) tree automaton~$A$ %(in particular the TA computing on the AST of a given CF-grammar)
a $\SWVPA$ $A$ such that $A(\lin(\tau))$ is the weight the AST $\tau$ of $G$.

>>>>

\begin{example}\label{ex:SWT}
Let us develop the example of comparison between music played by a performer,
represented as a sequence $s \in \Sigma^*$ of events in the MIDI alphabet $\Sigma$,
and  a music score represented as a sequence $t \in \Delta^*$
in the CWMN alphabet $\Delta$. We build a small weighted transducer model
 with two states $q_0$ and $q_1$ that
calculates the distance between $s$ and $t$.

@@

\begin{example}
Consider once more the scores of Exemple~\ref{ex:running}.
$\Deltai$ represents the set of music events of the form $<n, t>$
where $n$ is either a note (e.g. A4), the continuation symbol $-$, or a
rest symbol R. $\Deltac$ and  $\Deltar$ are used to encode the score structure:
Let $\Deltac = \{[, <]\}$ and  $\Deltar = \{], >]\}$ where $[$ (rep. $]$)
and $<$ (resp. $>$) correspond to start/end of measures, and start/end of
beams. Then  \includegraphics[scale=0.20]{pictures/score5.png}
is encoded with the nested word $[0, \frac{3}{4}, \frac{7}{8}>][<1, \frac{4}{3}, \frac{5}{3}]$
\endex
\end{example}

@@

@@

\item[--] a \SWVPA $A$ over $\Delta$, $\Semiring$, and $\bar\Phi$,
defining a measure $A: \Delta^* \to \Semiring$,
%a \SWVPA $A$ over $\Delta$, and $\Semiring$, defining a series of nested words
%      $A : \Delta^* \to \Semiring$,
\item[--] an input word $s \in \Sigma^*$.
\end{description}
%
For all $u \in \Sigma^*$ and $t \in \Delta^*$, let
\(d(u, t) = T\bigl(u, t|_\Deltai \bigr)\),
where $t|_\Deltai \in \Deltai^*$ is the projection of $t$ onto~$\Deltai$,
obtained from $t$ by removing all symbols in $\Delta \setminus \Deltai$.
%\end{equation}
%
\noindent
\emph{Symbolic weighted parsing} is the problem,
given the above input,
to find $t \in \Delta^*$ %nested word $t \in \Delta$
minimizing \( d(s, t) \otimes A(t)\)
\wrt $\leq_\oplus$,
\ie s.t. %such that: %Hence, it is the problem of finding
%
\begin{equation}\label{eq:distance-lang}
d(s, t) \otimes A(t) = \displaystyle\bigoplus_{t' \in \Delta^*} d(s, t') \otimes A(t')
\end{equation}
%
Following the terminology of~\cite{Mohri03EDWA},
$\SW$-parsing is the problem of computing
the distance~\eqref{eq:distance-lang} between the input~$s$ and the output weighted language of~$A$,
and returning a witness $t$.

\begin{example}[Symbolic Weighted Parsing and the transcription problem]
Applied to the music transcription problem, the above formalism
is interpreted as follows:
\begin{enumerate}

  \item The \SWT $T$ evaluates a ``fitness measure''  that expresses
    a correspondance between a performance and its notation. See Example~\ref{ex:SWT}.
\item The \SWVPA $A$ expresses a cost related to the music notation. One possibility
 is to relate this cost to the structural complexity: given a set of equivalent
 representations, we aim at choosing the simpler one. For instance
 \includegraphics[scale=0.20]{pictures/score4.png} should be favored
 on \includegraphics[scale=0.20]{pictures/score5.png}, although both yield
 a same linearization. Costs  can be expressed
 in the grammar with weight associated to each production rule. See Example~\ref{rules-division}.

 \end{enumerate}
Therefore, the framework, applied to the transcription problem, allows to find an optimal
solution that considers both the fitness of the result to the input, and the complexity of the former.
\end{example}
%
%\subsection{Computation}
%
\begin{proposition}
The problem of Symbolic Weighted  parsing
can be solved in PTIME in the size of the input \SWT $T$, \SWVPA $A$
and input word $s$,
and the computation time of the functions and operators of the label theory.
\end{proposition}
%
\begin{proof} (sketch)
We follow a \emph{Bar-Hillel} construction, for parsing by intersection.
%
\noindent
Let us first extend the \SWT $T$ over $\Sigma$, $\Deltai$
into a \SWT $T'$ over $\Sigma$ and $\Delta$
(and the same semiring and label theory $\Semiring$ and $\bar\Phi$),
such that for all $u \in \Sigma^*$, and $t \in {\Delta}^*$,
$T'(u, u) = T(u, t|_{\Deltai})$.
%
The transducer $T'$ simply skips every symbol
$b \in {\Delta} \setminus \Deltai$,
by the addition to $T$,
of new transitions of the form $\wei_{01}(q, \varepsilon, b, q')$.
%
\noindent
Then, using Corolary~\ref{cor:epsilon},
we construct from the input word $s \in \Sigma^*$ and $T'$
a \SWA $B_{s, T'}$,
such that for all $t \in \Delta^*$, $B_{s, T'}(t) = d(s, t)$.
%
%This automaton is such that for all $t \in \Delta^*$,
%\[
%   A_{T', s}\bigl(\lin(t)\bigr)
% = T'\bigl(s, \lin(t)\bigr)
% = T'\bigl(s, \lin(t)|_{\Deltai}\bigr)
% = d(s, t).
%\]
%
\noindent
Next, %we convert the input \SWTA $A$ over $\Delta$
%into a \SWVPA $A'$ over $\hat\Delta$, using Proposition~\ref{lem:SWTA}, and
we compute the \SWVPA $B_{s, T'} \otimes A$,
using Proposition~\ref{prop:SWVPA-product}.
%
\noindent
It remains to compute a best nested-word $t \in {\Delta}^*$
using the best-search procedure of Proposition~\ref{th:best-search}.
%and convert it into a best tree in $\T_\Delta$ in order to solve SW parsing
%for~$T$, $A$ and~$s$.
%\qed
\end{proof}
The $\SW$-parsing generalizes
the problem of searching the best derivation (AST) of a weighted CF-grammar $G$
that yields a given input word $w$.
The latter problem, sometimes called~\emph{weighted parsing},
(see \eg~\cite{Goodman99SemiringParsing}
 and~\cite{MorbitzVogler19weighted-parsing} for general weighted parsing frameworks)
corresponds to $\SW$-parsing in the case of finite alphabets,
a transducer $T$ computing the identity and some \SWVPA~$A$
obtained from $G$. %the weighted CF grammar.
%
%Indeed, it corresponds to the case where $T$
%accepts only the pairs $\<s, t>$ such that
%$s$ is the projection of $t$ on $\Deltai$.
%This can be done with a single state $q$ and
%with transition rules of the form:
%\begin{description}
%\item[] $\wei(q, \varepsilon, a, q) = \one$ for all $a \in \Deltac \cup \Deltar$,
%\item[] $\wei(q, a, a, q) = \one$ for all $a \in \Deltai$,
%\item[] $\wei(q, a, b, q) = \zero$ for all $a, b \in \Deltai$, $a \neq b$.
%\end{description}
%
Indeed, the \emph{depth-first} traversal of an AST $\tau$
yields a well-parenthesised word $\lin(\tau)$ over an alphabet
$\Delta = \Deltai \uplus \Deltac \uplus \Deltar$,
assuming \eg that $\Deltai$ contains the symbols labelling the leaves of $\tau$ (symbols of rank $0$),
and $\Deltac$ and $\Deltar$ contain respectively one left and right parenthesis
$\ccall{b}$ and $\creturn{b}$ for each symbol $b$ labelling inner nodes of $\tau$ (symbols of rank $>0$).
%
%With this representation, the projection $\lin(t)|_\Deltai$ is the
%sequence of leaves of $\tau$. %, enumerated in a \emph{dfs}-traversal.
%
We show in Appendix~\ref{sec:trees} how to construct
%($\SW$) tree automaton~$A$ %(in particular the TA computing on the AST of a given CF-grammar)
a $\SWVPA$ $A$ such that $A(\lin(\tau))$ is the weight the AST $\tau$ of $G$.
